The impacts of park access on health outcomes: a spatial comprehensive approach

### 2019-01-02 manuscript.Rnw
Edit notation in methodology section

\maketitle
<<setup, cache=FALSE, echo=FALSE, message=FALSE>>=
library(tidyverse)
library(geojsonio)
library(rgdal)
library(rgeos)
library(maptools)
library(spdep)
library(sf)
library(VGAM)
library(qwraps2)
library(knitr)
library(kableExtra)
opts_chunk$set(cache=TRUE)
library(leaflet)

@@

\section{Choice-based Accessibility}

Consider that an individual is choosing a park for a recreation activity.
According to basic choice theory \citep{McFadden1974}, the probability of choosing
park $p$ from the set of all regional parks $J$ is:

$$ P(p | V_p) = \frac{\exp(V_p)}{\sum_{j \in J}\exp(V_j)}$$
where parks are differentiated from each other by their relative measurable
utilities $V$. In principle, $V$ may include any measurable attributes of either
the choice maker or the park. In this study we use a linear formulation of

$$V_{ij} = size_j\beta_s + distance_{ij}\beta_d$$
incorporating the size of the park in acres and the distance of the park from
the census tract $i$ in miles. The coefficients $\beta$ are typically estimated from
surveys, though in the absence of a survey we apply a maximum likelihood
technique described below.

A key theoretical understanding of random utility choice models is that the
consumer surplus of the choice set can be otained as the log-sum of the
denominator of the choice probability equation. In plainer terms, the *total value*
of an individual's park accessibility is defined to be:

$$CS_i = \ln\left({\sum_{j \in J}\exp(V_{ij})}\right)$$

There are several advantages to a log-sum defined metric relative to
buffer-based accessibility metrics more commonly found in the literature. First,
all individuals are defined as having some access to all parks, rather than an
arbitrary limit of 1/2 mile or so. This allows for the fact that some people are
more or less sensitive to distances, and that distance is a continuous, and not
a binary, phenomenon. Second, the random utility formulation allows the
researcher to include any attribute of the park; in this case, we consider the
size of the park as an element of accessibility. This suggests that not all
parks are equal, and that a large park such as New York City's Central Park may
provide health and activity benefits over a much larger area than a smaller
community square.

Note that this formulation is an extension of the gravity-based accessibility
statistics, in the same way that the gravity model is a specific case of
a destination-choice model \citep{Daly1982}. The extension

@@

We predict the obesity rate in a census tract as a function of the tract's
sociodemographic characteristics and choice-based accessibility to parks with a
semi-log specification:
\begin{equation}\label{eq:themodel}
\log(\mathit{Obesity})_i = f(\vect{X}_i, A_i, g(W, \mathit(Obesity), \vect{X}))
\end{equation}
<<base_formula>>=
base_formula <- formula( 
  ~ log(density) + 
    fulltime + college + single + 
    Pct0to17 + Pct18to29 + Pct65plus + # need to leave out a category for collinearity  
    black + asian + hispanic + other)
@
where $\vect{T}_i$ is a vector of tract-level attributes, $A_i$ is a choice-based
accessibility measure for the tract, and $g(W, \mathit(Obesity), \vect{T})$
represents the impact of spatial neighborhood effects on the outcome.
Using tracts as the unit of observation in the model allows us to consider the
availability of parks at a metropolitan level, but it also introduces the problem
of spatial correlation of unobserved errors and spatial dependence in the data
generating process \citep{Cliff1970, Anselin1980}. Spatial correlation occurs
when spatially-distributed unobservable or missing variables (school quality,
neighborhood prestige, etc.) influence the modeled outcomes. Spatial dependence
occurs when the outcome {\em depends} on the outcome in neighboring areas; for
example, seeing neighbors exercising may encourage exercise. The relaxed spatial
Durbin model (SDM) proposed by \citet{Burridge1981} controls for both processes:
\begin{equation}\label{eq:sdm}
  \vect{y}= \rho W\vect{y} + X\vect{\beta} + WX\vect{\gamma} + \vect{\epsilon}
\end{equation}
where $\rho$ is the estimated spatial correlation between the outcome variables
$y_i, y_j, \ldots  \in \vect{y}$, $W$ is a matrix specifying the spatial
relationship between observations $i$ and $j$ in the dataset,
$\vect{\beta}$ is a vector of estimable coefficients relating the attributes $X$
of an observation to its outcome, and $\vect{\gamma}$ is a vector of
estimated coefficients relating the attributes of an observation's {\em neighbors}
to its outcome.

@@

We estimate

@@

A likelihood ratio test reveals that the SEM is not preferred, so we use the SDM
only going forward.

>>>>

\maketitle
<<setup, cache=FALSE, echo=FALSE, message=FALSE>>=
library(tidyverse)
library(geojsonio)
library(rgdal)
library(rgeos)
library(maptools)
library(spdep)
library(sf)
library(VGAM)
library(qwraps2)
library(knitr)
library(kableExtra)
opts_chunk$set(cache=TRUE, message = FALSE, echo = FALSE)
library(leaflet)

@@

\subsection{Choice-based Accessibility}

Consider that an individual is choosing a park for a recreation activity.
If we apply random utility choice theory, specifically the multinomial
logit model \citep{McFadden1974}, the expected probability of individual 
$i$ choosing park $j$ from the set of all regional parks $P$ is
\begin{equation}\label{eq:mnl}
P(j | V_{ij}) = \frac{\exp(V_{ij})}{\sum_{p \in P}\exp(V_{ip})}
\end{equation}
where parks are differentiated from each other by their relative measurable
utilities $V_{ij}$. In principle, $V$ may include any measurable attributes of either
the choice maker or the park. In this study we use a linear formulation of
\begin{equation}\label{eq:v_sd}
V_{ij} = size_j\beta_s + distance_{ij}\beta_d
\end{equation}
incorporating the size of the park in acres and the distance of the park from
the census tract $i$ in miles. The coefficients $\vec{\beta}$ can be 
estimated from household surveys, though in the absence of a survey we 
may assert reasonable values.

A key theoretical understanding of random utility choice models is that the
consumer surplus, or the total value of all alternatives of the choice set, can 
be obtained as the log-sum of the denominator of the choice probability equation:
\begin{equation}\label{eq:cs}
CS_i = \ln\left({\sum_{p \in J}\exp(V_{ip})}\right)
\end{equation}
There are several advantages to a log-sum defined metric relative to
buffer-based accessibility metrics more commonly found in the literature. First,
all individuals are defined as having some access to all parks, rather than an
arbitrary cutoff asserted by the researcher. This allows for the fact that some 
people are more or less sensitive to distances, and that distance is a 
continuous, and not a binary, phenomenon. Second, the random utility 
formulation allows the researcher to include any attribute of the park; 
in this case, we consider the
size of the park as an element of accessibility. This suggests that not all
parks are equal, and that a large park such as Central Park on Manhattan may
provide health and activity benefits over a much larger area than a smaller
community square.

Note that this formulation is an extension of the gravity-based accessibility
statistics, in the same way that the gravity model is a specific case of
a destination-choice model \citep{Daly1982}. The extension allows us to consider
more attributes of the park or the indivdual beyond size and impedance. For example,
we can aggregate the number of tweets emanating from a park as a measure of the
park's attractiveness beyond its size, and the utility for the choice maker from 
Equation \ref{eq:v_sd} then becomes
\begin{equation}\label{eq:v_sdt}
V_{ij} = size_j\beta_s + distance_{ij}\beta_d + tweets_j\beta_t
\end{equation}

@@

We predict the obesity rate in a census tract as a function of the tract's
sociodemographic characteristics and choice-based accessibility to parks with a
semi-log specification:
\begin{equation}\label{eq:themodel}
\log(\mathit{Obesity})_i = f(\vect{x}_i, CS_i, g(W, \mathit{Obesity}, \vect{X}))
\end{equation}
<<base_formula>>=
base_formula <- formula( 
  ~ log(density) + 
    fulltime + college + single + 
    Pct0to17 + Pct18to29 + Pct65plus + # need to leave out a category for collinearity  
    black + asian + hispanic + other)
@
where $\vect{x}_i$ is a vector of tract-level attributes, $CS_i$ is a choice-based
accessibility measure for the tract computed from Equations \ref{eq:v_sd} or 
\ref{eq:v_sdt}, and $g(W, \mathit{Obesity}, T)$
represents the impact of spatial neighborhood effects on the outcome.
Using tracts as the unit of observation in the model allows us to consider the
availability of parks at a metropolitan level, but it also introduces the problem
of spatial correlation of unobserved errors and spatial dependence in the data
generating process \citep{Cliff1970, Anselin1980}. Spatial correlation occurs
when spatially-distributed unobservable or missing variables (school quality,
neighborhood prestige, etc.) influence the modeled outcomes. Spatial dependence
occurs when the outcome {\em depends} on the outcome in neighboring areas; for
example, seeing neighbors exercising may encourage exercise. The relaxed spatial
Durbin model (SDM) proposed by \citet{Burridge1981} controls for both processes:
\begin{equation}\label{eq:sdm}
  \vect{y}= \rho W\vect{y} + X\vect{\beta} + WX\vect{\gamma} + \vect{\epsilon}
\end{equation}
where $\rho$ is the estimated spatial correlation between the outcome variables
$y_i, y_j, \ldots  \in \vect{y}$, $W$ is a matrix specifying the spatial
relationship between observations $i$ and $j$ in the dataset,
$\vect{\beta}$ is a vector of estimable coefficients relating the attributes $X$
of an observation to its outcome, and $\vect{\gamma}$ is a vector of
estimated coefficients relating the attributes of an observation's {\em neighbors}
to its outcome.

@@

@@

We estimated the A likelihood ratio test rejects that the SEM is equivalent, so we use the SDM
only going forward.
